For submission to the Journal of Integrated Pest Management

### 2021-03-02 ofv_fife_2021.Rmd
Made minor change, waiting on Dr. Iriarte to send info on her methods
Accepted Dr. Martini's suggestions to the paper

### First report of the *Brevipalpus*-transmitted (Trombidiformes: Tenuipalpidae) virus, *Orchid fleck dichorhavirus* (Mononegavirales: Rhabdoviridae) infecting various Nolinoideae in the Florida panhandle

@@

^2^ University of Florida, Department of Entomology and Nematology,Tropical Research and Education Center, Homestead FL 33031

@@

### Abstract
Flat mites from the genus *Brevipalpus* Donnadieu (Trombidiformes: Tenuipalpidae) are the only known vectors of dichorhaviruses and transmit *Orchid fleck dichorhavirus* (OFV) in a persistent propagative manner. OFV is the type species for the genus *Dichorhavirus* and infects 50+ spp. of Orchidaceae, Nolinoideae, and *Citrus*.
During June 2020, we observed chlorotic ringspot symptoms on Giant Lilyturf (*Liriope* spp., cv. ‘Gigantea’) in the landscape of Leon County, Florida. We tested plant samples for a wide range of pathogens and ultimately identified the pathogen as *Orchid fleck dichorhavirus*. We took further samples from other *Liriope* spp., *Ophiopogon* spp., and *Aspidistra elatior* Blume (Asparagales: Nolinoideae) in Leon and Alachua county. All symptomatic plants tested positive for OFV. We recovered *Brevipalpus* mites from OFV-infected plants, which belong to *Brevipalpus californicus* sensu lato, associated with Nolinoideae.
Florida has various pest species of *Brevipalpus* and a diverse array of susceptible plant species in the landscape, both native and introduced. We suggest that OFV has a wide distribution in Florida which will continue to spread if unchecked, representing a potential threat for many plants of economic importance for Florida.

### Keywords:
false spider mite, flat mite, Brevipalpus-transmitted viruses, Liriope, Ruscaceae, Asparagaceae, orchid, Orchidaceae, pests, ornamental plants

@@

OFV-infected plants exhibit various symptoms dependent on the variety of plant infected as well as the strain of the virus [@Kubo2009b], but symptoms typically appear as chlorotic flecks, which ultimately coalesce into larger spots or ringspot patterns (see *`r citef("fig_1")`*).
<!-- These lesions are thought to be associated with the feeding sites of the mite vector [@cite]. Also, is it probably a sign of RNAi plant defenses: see Hull Plant Virology -->

@@

OFV naturally infects 50+ spp of Orchidaceae [@Kitajima2010; @Peng2013], some Nolinoideae [@Mei2016, @Dietzgen2018b], and *Citrus* [@BeltranBeltran2020; @GarciaEscamilla2018 @Roy2020]. Mechanical transmission of OFV is possible under lab conditions to various *Chenopodiaceae*, *Aizoaceae*, *Fabaceae*, and *Solanaceae* [@Chang1976; @Kondo2003; @Peng2013].

On June 2020, chlorotic ringspot symptoms were observed on the liriopogon Giant Lilyturf *Liriope* spp., cv. 'Gigantea' in the landscape of Leon County, Florida (see *`r citef("fig_1")`*). Liropogons belong to the plant family Asparagaceae, subfamily Nolinoideae, which includes various monocotyledonous lilliod plants [@Chase2009]. Liriopogons are native to southeastern Asia, encompassing the closely-related genera *Liriope* and *Ophiopogon*, which are characterized by evergreen, long grass-like leaves, and erect flowers which produce black, purple or blue, berry-like fruits, depending on the cultivar [@Lattier2014; @Fantz2015]. These characteristics have given rise to various common names, including: monkey grass, Aztec grass, lilyturf, mondo grass, etc. (see *`r citet("tab_1")`*). The genus *Liriope* contains eight species, four of which are commonly cultivated, while the genus *Ophiopogon* consists of about 65 species, with only a handful are available for horticulture [@Lattier2014; @Nesom2010]. Liripogons are considered the most important ground cover sold by the nursery industry in southeastern US [@Mcharo2003].

Plant leaf samples were initially tested at the Plant Disease Diagnostic Clinic at the North Florida Research and Education Center (NFREC) in Quincy, FL. The plant samples were tested for a wide range of pathogens, including **[Ask Dr. Iriarte]** but the results were inconclusive. The infected materials were subsequently sent to the Florida Department of Agriculture and Consumer Services (FDACS). The presence of *Orchid fleck dichorhavirus* was confirmed via generic one step conventional RT-PCR with R2-Dicho-GF and R2-Dicho-GR primers, amplifying ~800 nt of L-gene (RNA2) amplicon from a leaf sample of infected *Liriope* [@Kubo2009b]. RT-qPCR confirmed OFV and the RT-PCR product was sequenced using Sanger sequencing and High Throughput Sequencing [@Kondo2006]. Sanger sequencing of RT-PCR amplicons shared 98 \% nucleotide identity with OFV strains which infect orchids (GenBank Accession numbers: AB244418 and MK522807) [@Kondo2006; @Kondo2014; @Cook2019]. Further samples were taken from various symptomatic cultivars of *Liriope* spp., *Ophiopogon* spp., as well as *Aspidistra elatior* Blume (Asparagales: Nolinoideae) during subsequent visits to the initial site of collection as well as other locations in Leon county. Samples were tested via RT-PCR at the NFREC (see *`r citet("tab_1")`*) to confirm the presence of OFV.

Further surveys of Nolinoideae in Florida have revealed more sites with symptomatic plants in both Leon and Alachua counties. In addition to the discovery of the OFV-infected plants, our lab at the NFREC collected mites from symptomatic plants in Leon county which were observed with phase contrast microscopy. We encountered both eriophyoid mites and flat mites which were identified as *Brevipalpus californicus* (Banks) sensu lato, an identity which was confirmed by the FDACS via Differential Interference Contrast (DIC) microscopy. *Brevipalpus* mites have been previously associated with OFV [@BeltranBeltran2020; @GarciaEscamilla2018; @Dietzgen2018b] and similar diseases [@Kitajima2010] and are known to feed on a large variety of economically-important plants [@Childers2003a; @Akyazi2017]. Unfortunately, the *Brevipalpus* mite species complex is known to contain cryptic species [@Childers2011] whose identification can be improved with molecular methods as well as more advanced microscopy techniques, such as low-temperature scanning electron microscopy (LT-SEM) [@Skoracka2015; @Leon2010]. With that in mind, we sent additional samples of the mites to the USDA-ARS in Beltsville to observe the mites with LT-SEM techniques, which agreed with both prior identifications of *B. californicus* s.l. (see *`r citef("fig_2")`*).

The earliest mention of a virus which may have been OFV is @Ko1985, who detected nuclear inclusions caused by an undescribed bacilliform rhabdovirus. @Ko1985 described the spoke-wheel configurations typically associated with OFV (see [@Chang1976]), in *Brassia* orchids, but unfortunately made no mention of mites or further investigations of this virus. The first certain report of OFV in the US was made by @Bratsch2015, who confirmed the presence of OFV in *Phalaenopsis* hybrids in the US, using TEM of ultrathin sections of plant tissue as well as molecular sequence analysis. In addition, @Bratsch2015 detected *Brevipalpus* mites and their exoskeletons associated with OFV-infected plants. It appears that @Bratsch2015 did not make a conclusive species identification, but did cite @Kondo2003, suggesting that they suspected *B. californicus* as their vector.

@@

Detecting OFV in Florida represents a concern for horticulturalists who grow susceptible plants of economic importance, including orchids, *Liriope*, *Ophiopogon*, or other susceptible Asparagales spp. which are commonly used in landscaping. Florida is also home to a plethora of native and naturalized orchid species, many of which are threatened, including the famous Ghost Orchid, *Dendrophylax lindenii* (Lindl.) Benth. ex Rolfe. Furthermore, OFV represents an obstacle to overcome for the burgeoning interest in cultivating *Vanilla* in southern Florida [@Chambers2019]. Lastly, some OFV isolates are known to be involved with nuclear types of Citrus Leprosis [@Roy2015a], which may be a cause for concern for the citrus industry. In fact, @Kitajima2011a found that the Citrus Leprosis virus (CiLV) which previously affected Florida citrus was a nuclear type of citrus leprosis, which are closely related to OFV strains [@Roy2013a]. *B. californicus* and *B. yothersi* are both known vectors of Dichorhaviruses (OFV) and Cileviruses (Citrus Leprosis) [@BeltranBeltran2020; @Kondo2003; @Knorr1968] and *B. obovatus* is a suspected vector as well [@Childers2003a]. All three mite species/complexes are present in Florida [@Childers2003a; @Akyazi2017] (see *`r citef("fig_3")`*), therefore it is critical to monitor the spread of Orchid Fleck Virus and its mite vector(s) to determine what risk this virus represents for plants in Florida and the surrounding regions.

### Control Options

*Galendromus helveolus* (Acari: Phytoseiidae) have been reported to feed on eggs, larvae and nymphs of *B. californicus* [@Chen2006].

*Metarhizium anisopliae* var. *acridum* has been shown to be pathogenic to *B. phoenicis* [@Magalhaes2005], we have encountered similar fungal infections

>>>>

### First report of the *Brevipalpus*-transmitted (Trombidiformes: Tenuipalpidae) virus, *Orchid fleck dichorhavirus* (Mononegavirales: Rhabdoviridae) infecting ornamental Nolinoideae in the Florida panhandle

@@

^2^ University of Florida, Department of Entomology and Nematology, Tropical Research and Education Center, Homestead FL 33031

@@

### Abstract
Flat mites from the genus *Brevipalpus* Donnadieu (Trombidiformes: Tenuipalpidae) are the only known vectors of dichorhaviruses and transmit *Orchid fleck dichorhavirus* (OFV) in a persistent propagative manner. OFV is the type species for the genus *Dichorhavirus* and infects 50+ spp. of Orchidaceae, Nolinoideae, and *Citrus*.
During June 2020, we observed chlorotic ringspot symptoms on Giant Lilyturf (*Liriope* spp., cv. ‘Gigantea’) in the landscape of Leon County, Florida. We tested plant samples for a wide range of pathogens and ultimately identified the pathogen as *Orchid fleck dichorhavirus*. We took further samples from other *Liriope* spp., *Ophiopogon* spp., and *Aspidistra elatior* Blume (Asparagales: Nolinoideae) in Leon and Alachua county. All symptomatic plants tested positive for OFV. We recovered mites from OFV-infected plants, which belonged to *Brevipalpus californicus* sensu lato.
Florida has various mite species of *Brevipalpus* and a diverse array of susceptible plant species in the landscape, both native and introduced. We suggest that OFV already has a wide distribution in Florida which will continue to spread if unchecked, representing a potential threat for many plants of economic importance for Florida.

### Keywords:
false spider mite, flat mite, *Brevipalpus*-transmitted viruses, *Liriope*, Ruscaceae, Asparagaceae, orchid, Orchidaceae, pests, ornamental plants

@@

OFV-infected plants exhibit various symptoms dependent on the variety of plant infected as well as the strain of the virus [@Kubo2009b], but symptoms typically appear as chlorotic flecks, which ultimately coalesce into larger spots or ringspot patterns (*`r citef("fig_1")`*).
<!-- These lesions are thought to be associated with the feeding sites of the mite vector [@cite]. Also, is it probably a sign of RNAi plant defenses: see Hull Plant Virology -->

@@

OFV naturally infects 50+ spp of Orchidaceae [@Kitajima2010; @Peng2013], some Nolinoideae [@Mei2016, @Dietzgen2018b], and *Citrus* where it causes citrus leprosis-like symptoms [@BeltranBeltran2020; @GarciaEscamilla2018 @Roy2020]. Mechanical transmission of OFV is possible under lab conditions to various *Chenopodiaceae*, *Aizoaceae*, *Fabaceae*, and *Solanaceae* [@Chang1976; @Kondo2003; @Peng2013].

On June 2020, chlorotic ringspot symptoms were observed on the liriopogon Giant Lilyturf *Liriope* spp., cv. 'Gigantea' in the landscape of Leon County, Florida (*`r citef("fig_1")`*). Liropogons belong to the plant family Asparagaceae, subfamily Nolinoideae, which includes various monocotyledonous lilliod plants [@Chase2009]. Liriopogons are native to southeastern Asia, encompassing the closely-related genera *Liriope* and *Ophiopogon*, which are characterized by evergreen, long grass-like leaves, and erect flowers which produce black, purple or blue, berry-like fruits, depending on the cultivar [@Lattier2014; @Fantz2015]. These characteristics have given rise to various common names, including: monkey grass, Aztec grass, lilyturf, mondo grass, etc. (*`r citet("tab_1")`*). The genus *Liriope* contains eight species, four of which are commonly cultivated, while the genus *Ophiopogon* consists of about 65 species, with only a handful are available for horticulture [@Lattier2014; @Nesom2010]. Liripogons are considered the most important ground cover sold by the nursery industry in southeastern US [@Mcharo2003].

Plant leaf samples were initially tested at the Plant Disease Diagnostic Clinic at the North Florida Research and Education Center (NFREC) in Quincy, FL. The plant samples were tested for a wide range of pathogens, including **[Waiting on Dr. Iriarte's response]** but the results were inconclusive. The infected materials were subsequently sent to the Florida Department of Agriculture and Consumer Services (FDACS). The presence of *Orchid fleck dichorhavirus* was confirmed via generic one step conventional RT-PCR with R2-Dicho-GF and R2-Dicho-GR primers, amplifying ~800 nt of L-gene (RNA2) amplicon from a leaf sample of infected *Liriope* [@Kubo2009b]. RT-qPCR confirmed OFV and the RT-PCR product was sequenced using Sanger sequencing and High Throughput Sequencing [@Kondo2006]. Sanger sequencing of RT-PCR amplicons shared 98 \% nucleotide identity with OFV strains which infect orchids (GenBank Accession numbers: AB244418 and MK522807) [@Kondo2006; @Kondo2014; @Cook2019]. Further samples were taken from various symptomatic cultivars of *Liriope* spp., *Ophiopogon* spp., as well as *Aspidistra elatior* Blume (Asparagales: Nolinoideae) during subsequent visits to the initial site of collection as well as other locations in Leon county. Samples were tested via RT-PCR at the NFREC (*`r citet("tab_1")`*) to confirm the presence of OFV.

Further surveys of Nolinoideae in Florida have revealed more sites with symptomatic plants in both Leon and Alachua counties. In addition to the discovery of the OFV-infected plants, we collected mites from symptomatic plants in Leon county which were observed with phase contrast microscopy. We encountered both eriophyoid mites and flat mites which were identified as *Brevipalpus californicus* (Banks) sensu lato, an identity which was confirmed by the FDACS via Differential Interference Contrast (DIC) microscopy. *Brevipalpus* mites have been previously associated with OFV [@BeltranBeltran2020; @GarciaEscamilla2018; @Dietzgen2018b] and similar diseases [@Kitajima2010] and are known to feed on a large variety of economically-important plants [@Childers2003a; @Akyazi2017]. Unfortunately, the *Brevipalpus* mite species complex is known to contain cryptic species [@Childers2011] whose identification can be improved with molecular methods as well as more advanced microscopy techniques, such as low-temperature scanning electron microscopy (LT-SEM) [@Skoracka2015; @Leon2010]. With that in mind, we sent additional samples of the mites to the USDA-ARS in Beltsville to observe the mites with LT-SEM techniques, which agreed with both prior identifications of *B. californicus* s.l. (*`r citef("fig_2")`*).

The earliest mention of a virus which may have been OFV is @Ko1985, who detected nuclear inclusions caused by an undescribed bacilliform rhabdovirus. @Ko1985 described the spoke-wheel configurations typically associated with OFV [@Chang1976], in *Brassia* orchids, but unfortunately made no mention of mites or further investigations of this virus. The first certain report of OFV in the US was made by @Bratsch2015, who confirmed the presence of OFV in *Phalaenopsis* hybrids in the US, using TEM of ultrathin sections of plant tissue as well as molecular sequence analysis. In addition, @Bratsch2015 detected *Brevipalpus* mites and their exoskeletons associated with OFV-infected plants. It appears that @Bratsch2015 did not make a conclusive species identification, but did cite @Kondo2003, suggesting that they suspected *B. californicus* as their vector.

@@

Detecting OFV in Florida represents a concern for horticulturalists who grow susceptible plants of economic importance, including orchids, *Liriope*, *Ophiopogon*, or other susceptible Asparagales spp. which are commonly used in landscaping. Florida is also home to a plethora of native and naturalized orchid species, many of which are threatened, including the famous Ghost Orchid, *Dendrophylax lindenii* (Lindl.) Benth. ex Rolfe. Furthermore, OFV represents an obstacle to overcome for the burgeoning interest in cultivating *Vanilla* in southern Florida [@Chambers2019]. Lastly, some OFV isolates are known to be involved with nuclear types of citrus leprosis [@Roy2015a], which may be a cause for concern for the citrus industry. In fact, @Kitajima2011a found that the Citrus Leprosis virus (CiLV) which previously affected Florida citrus was a nuclear type of citrus leprosis, which are closely related to OFV strains [@Roy2013a]. *B. californicus* and *B. yothersi* are both known vectors of Dichorhaviruses (OFV) and Cileviruses (Citrus Leprosis) [@BeltranBeltran2020; @Kondo2003; @Knorr1968] and *B. obovatus* is a suspected vector as well [@Childers2003a]. All three mite species/complexes are present in Florida [@Childers2003a; @Akyazi2017] (*`r citef("fig_3")`*); therefore, it is critical to monitor the spread of Orchid Fleck Virus and its mite vector(s) to determine what risk this virus represents for plants in Florida and the surrounding regions.

### Brief Section on Control Options

The predatory mite *Galendromus helveolus* (Acari: Phytoseiidae) have been reported to feed on eggs, larvae and nymphs of *B. californicus* [@Chen2006].

The entomopathogenic fungus *Metarhizium anisopliae* var. *acridum* has been shown to be pathogenic to *B. phoenicis* [@Magalhaes2005], we have encountered similar fungal infections

<div id="refs"></div>
